Machine learning and essentialism

### 2023-02-06 ART_Sekrst/Sekrst.tex
Final commit before sending to the editors for corrections

\begin{artengenv2auth}{Kristina Šekrst, Sandro Skansi}
	{Machine learning and essentialism}
	{Machine learning and essentialism}
	{Machine learning and essentialism}
	{\textsuperscript{1}Departments of Philosophy and Mathematics, Carnegie Mellon University\\
		\textsuperscript{2}Copernicus Center for Interdisciplinary Studies, Jagiellonian University}
	{Machine learning and essentialism have been connected in the past by various researchers, in order to state that the main paradigm in machine learning processes is equivalent to choosing the ``essential'' attributes for the machine to search for. Our goal in this paper is to show that there are connections between machine learning and essentialism, but only for some kinds of machine learning, and often not including deep learning methods. Similarity-based approaches, more connected to the overall prototype theory, spanning from psychology and linguistics, seem more suited for pattern recognition and complex deep-learning issues, while for classification problems, mostly for unsupervised learning, essentialism seems like the best choice. In order to illustrate the difference better, we will connect both paths to their sources in other disciplines and see how human psychology influences our decision in machine-learning modeling as well. This leads to a~philosophically very interesting consequence: even in the setting of supervised machine learning, essences are not present in data, but in targets, which in turn means that the categories which purport to be essences are in fact human-made, and hand-coded in the targets. The success of machine learning, therefore, does not give any substantial evidence for the independent existence of essential properties. Our stance here is to state that neither the existence nor the lack of ``essential'' properties in machine learning can lead to metaphysical, i.e., ontological claims.}
	{essentialism, machine learning, accidental properties, similarity-based approach, pattern recognition, modal necessity.}
	{%
		{\flushright\subbold{Kristina Šekrst}\\\subsubsectit\small{University of Zagreb}\par}%
		{\flushright\subbold{Sandro Skansi}\\\subsubsectit\small{University of Zagreb}\par}%
	}

@@

\section{Essential and accidental properties: introduction}
The purpose of this paper is to show that the existence of essential-like features in machine learning, or the lack of them, cannot provide an ontological commitment.\footnote{The authors would like to thank the anonymous reviewers for their detailed analyses and insights.} Researchers have connected machine-learning practices with essentialist and anti-essentialist stances, but we feel that such claims ignore that both ``essentialist'' and ``anti-essentialist'' paradigms in machine learning are both influenced by human psychology and have no real consequence on the verification of whether there \textit{are} essential properties in nature or not.\footnote{It is necessary to distinguish between ontological commitments regarding nature and ontological commitments that are necessary in every AI system (we call this difference an ontological gap). The former are the subject of this article, and the latter were analyzed by Krzanowski \& Polak
%\label{ref:RNDU3VddFbPO5}(2022a; 2022b).
\parencites*[][]{krzanowski_ontology_2022}[][]{krzanowski_meta-ontology_2022}.%
}

@@

\section{Philosophical essentialism}
An \textit{essential} property of an object is a~property that an object must have, while an \textit{accidental} one is the one the object happens could have, but that it could lack. That is, in modal terms,\footnote{Standard modal characterizations have been disseminated with the works of Ruth Barcan Marcus and Saul Kripke. Kripke's work on semantics is taking the truth of a~formula relative to a~possible world, since its truth value depends on what is true in accessible world. See Barcan Marcus
%\label{ref:RNDktVIAPuvSw}(1993)
\parencite*[][]{marcus_modalities_1993} %
 for a~modality synthesis and Kripke 
%\label{ref:RND50iCIBiyUn}(1972)
\parencite*[][]{kripke_naming_1972} %
 for Kripke semantics.} we are talking about \textit{necessity} and \textit{possibility},\footnote{There are, of course, differences between logical and (meta)physical possibilities. Something might not be a~logical contradiction, but still be (meta)physically impossible, i.e., not conforming to the laws of nature, for example, a~man travelling faster than the speed of light. The exact details of such classifications, especially between physical and metaphysical possibilities, are a~matter of debate.} respectively 
%\label{ref:RNDcC0OGIZEmV}(Robertson Ishii and Atkins, 2020):
\parencite[][]{robertson_ishii_essential_2020}: %
 an object is going to possess the essential property in all possible worlds, but for an accidental one, there is a~possible world in which an object lacks such a~property. \textit{Essentialism} is a~standpoint in which (at least) some objects have (at least some) essential properties 
%\label{ref:RNDS25JSRGYR9}(Robertson Ishii and Atkins, 2020).
\parencite[][]{robertson_ishii_essential_2020}. %
 For example, an essential property of Socrates is to have originated from his parents but is not essential for him to have brown hair. An essential property of a~dog is certainly not brown hair since there are dogs of other colors. In philosophy, some essential properties are not a~matter of much debate. For example, a~dog had to have some biological origin. \textit{Canis canis} is also a~dog. But in order to pursue the matter further, there might be various objections to candidates for essential purposes.\footnote{A~concise description of the debate is provided by Cartwright 
%\label{ref:RNDsX991rAklO}(1968, p.615):
\parencite*[][p.615]{cartwright_remarks_1968}: %
 ``What are the essential attributes of, say, Dancer's Image? No doubt it will be counted essential that he is a~horse and accidental that he was disqualified in this year's Kentucky Derby. But what of the attribute of being male, or of being a~thoroughbred, or of not being a~Clydesdale stallion? Here, I~suppose, essentialists may disagree. Indeed, a~reasonable essentialist might well take the position that these are hard cases that admit of no clear decision.}'' Often, a~dog is considered a~``wolf-like descendant'', where various breeds might not conform to this ideal, along with the notion of ``having an upturning tail''.

@@

\section{Machine-learning basics}
\textit{Machine learning}, as a~part of artificial intelligence and computer science, is a~field of approaches and methods that use data in order to improve their performance on some problems. Artificial intelligence can be seen as a~certain type of philosophical engineering
%\label{ref:RNDwNayypCdyS}(Skansi, 2018, p.vii):
\parencite[][p.vii]{skansi_introduction_2018}: %
 we want ``to build machines that can think, [\ldots] understand the meaning, act rationally, cope with uncertainty, [\ldots] handle and talk about objects''. In a~nutshell, we are replicating standard philosophical concepts. It is no wonder that philosophical concepts are deeply embedded in their methods as well but may seem hidden underneath technical layers.

@@

\section{Psychological essentialism}
Gelman
%\label{ref:RNDqWgJbFowwn}(2004, p.405)
\parencite*[][p.405]{gelman_psychological_2004} %
 describes how once children learn a~new fact about one member of a~category, they generalize the fact to other members of that category, even if they look substantially different. By four years of age, children display subtlety and flexibility when they make category-based inductive inferences. For Gelman 
%\label{ref:RNDf5jZZXitzC}(2004, p.405),
\parencite*[][p.405]{gelman_psychological_2004}, %
 properties seem to be ``fixed at birth'', demonstrated by the following experiment. A~child might learn about a~newborn kangaroo that was switched at birth, and then went to live with goats. The child was then asked whether the animal would be good at hopping or climbing, or if would it have a~pouch or not. Turns out, preschool children typically reported that it would have been good at hopping and have a~pouch, something that seems \textit{inherent} to kangaroos even for children. Such an understanding seems to appear by about six years of age, and it might be as early as four years of age: the time when children reason about animals, plants, and social categories 
%\label{ref:RNDyFC6Tu5B0p}(Gelman, 2004, p.406).
\parencite[][p.406]{gelman_psychological_2004}.%

@@

\section{Features and prototypes}
As mentioned, \textit{features} tend to be measurable properties that are successful in discriminating and differentiating between different categories of data. For example, in face detection
%\label{ref:RND7NdRdpncD9}(Bishop, 2006, p.3),
\parencite[][p.3]{bishop_pattern_2006}, %
 we aim to find features that are not only fast to compute but also preserve useful discriminatory information enabling faces to be distinguished from non-faces. The study of feature selection finds its practical needs in machine learning, where a~learning algorithm constructs a~description of a~function from a~set of input/output instances through interaction with the world. Machine learning is more concerned with non-continuous features, while pattern recognition deals with continuous ones. It is not the same, for example, to classify something as a~dog or not, compared to finding a~face or another pattern or blob inside an image 
%\label{ref:RNDjCPrNP8QrO}(Liu and Motoda, 1998, p.2).
\parencite[][p.2]{liu_feature_1998}. %
 Liu and Motoda 
%\label{ref:RNDWGLVGs5aWN}(1998, p.2)
\parencite*[][p.2]{liu_feature_1998} %
 state that many forms of representations for machine-learning functions are available, including first-order logic, which is interesting from a~philosophical standpoint, or weighted networks, but they have focused on features since they are 1) \textit{primitive} 2) \textit{convenient} 3) \textit{independent} 4) \textit{widely used} 5) \textit{reasonably general}, i.e., powerful for many applications.

@@

\section{Essentialist paradigm(s) in machine learning}
It seems intuitive and obvious that supervised machine learning incorporates some kind of essentialism. That is, we are either given discrete or continuous features in datasets that are used for our predictions, usually whether something is a~member of a~class or not. But there are other kinds of machine learning, and we must not ignore the notion of unsupervised learning. We have already mentioned unsupervised learning, in which a~model tries to establish regularities, clusters, or patterns in previously unseen data. This can be compared to the process of human learning at an early age, in which a~human being tries to generalize the already acquired knowledge. Consider this, even if you are getting an unlabeled dataset of weird alien creatures, you will most certainly be able to connect similar ones together in groups or do classifications, even if you do not know what \textit{is} actually in the background of your dataset. We would like the computer to do the same. For example, if we trained our models on a~certain map, they might recognize landmasses, developed areas, forests, or wetlands and group them together, by finding similarities between them. In non-visual data, you might be presented with some numbers, say bank transfers, and you might connect the usual activity into groups, while the outliers might be suspicious.

@@

\section{Accidental properties in machine learning}
The processes and disciplines of \textit{feature selection} and \textit{extraction} show us that there is a~strong presupposition that something as essential as a~feature exists. There is no doubt that machine learning today is still enveloped in a~strong essentialist paradigm. In feature engineering,\footnote{For more details about feature engineering, see Zheng and Casari
%\label{ref:RNDzLhpHmjkOb}(2018).
\parencite*[][]{zheng_feature_2018}.%
} a~system automatically discovers representations needed for feature detection. For example, it finds close points (neighbors) in a~graph and clusters data around, say, percentages. If feature engineering is an essentialist stance, what kind of essentialism is it? It seems that is not maximal, but also not minimal, we would expect it to lie somewhere in between, judging by its success factor.

@@

\section{Machine learning or essentialism?}
Our previous conclusion might imply two separate things. Either there is an anti-cybernetic stance in which human learning that encompasses a~certain kind of innate essentialist knowledge is a~different process in machine learning, or that, for practical purposes, knowledge of essential properties is not a~necessary prerequisite for everyday classifications and predictions. The latter seems more intuitive. It does seem that a~similarity-based approach, mimicking the prototypical relationships found in psychological and linguistic research, may work well in various human and machine usages, along with a~combination of properties (features) together with their relations (cf. feature vectors). For some machine-learning tasks, pure essentialism, often a~binary or Boolean one, works best. We believe that essentialism and anti-essentialisms are not binary choices a~computer scientist or a~philosopher must make in order to describe how processes are being generated and run in machine learning paradigms today, but it is a~matter of choice \textit{for a~specific type of task}. There is no essentialism equated with machine learning, but there is both essentialism and anti-essentialism for specific tasks. For some classification tasks and simple pattern recognitions, essentialist features are often the best choice, and for others, systems will work better with combinations of these properties. For unsupervised learning and pattern recognition, prototypical systems, i.e., similarity-based approaches, perform better. A~philosophical take here is that, at least in machine learning, there is no ontological obligation towards either of these stances, but rightful usage for rightful tasks. The choice of your machine-learning system, and therefore, a~supervised or unsupervised approach, will depend on the type of task in question: \textit{what performs better}. It is just a~matter of technical performance that has no metaphysical consequences of the existence of essentialism or anti-essentialism.

@@

\end{artengenv2auth}

%\section{Bibliography}
%Aristotle, 2014. Categories. In: Jonathan Barnes, ed. \textit{The Complete Works of Aristotle: The Revised Oxford Translation, One-Volume Digital Edition}, Bollingen series, 6. print., with corr. Princeton, N.J: Princeton University Press. pp.25–70.
%
%Ben-Menahem, Y., 2006. \textit{Conventionalism[202F?]: From Poincare to Quine}. [online] Cambridge; New York: Cambridge University Press. Available at: {\textless}https://search.ebscohost.com/login.aspx?direct=true\&db=e000xww\&AN=529339\&lang=pl\&site=ehost-live{\textgreater} [Accessed 13 January 2023].
%
%Bishop, C.M., 2006. \textit{Pattern Recognition and Machine Learning}. Information science and statistics. New York: Springer.
%
%Cartwright, R.L., 1968. Some Remarks on Essentialism. \textit{The Journal of Philosophy}, [online] 65(20), pp.615–626. https://doi.org/10.2307/2024315.
%
%Cohen, M.F., 1968. Wittgenstein's anti-essentialism. \textit{Australasian Journal of Philosophy}, [online] 46(3), pp.210–224. https://doi.org/10.1080/00048406812341181.
%
%Descartes, R., 1641. \textit{Renati Des-Cartes Meditationes de prima philosophia, in qua Dei existentia et animae immortalitas demonstratur.} [online] Paris: Michael Soly. Available at: {\textless}https://gallica.bnf.fr/ark:/12148/btv1b86002964{\textgreater} [Accessed 25 August 2021].
%
%Descartes, R., 1991. Meditations on First Philosophy. Translated by J. Cottingham, R. Stoothoff and D. Murdoch. In: \textit{The Philosophical Writings of Descartes, vol. 2}. Cambridge: Cambridge University Press. pp.1–63.
%
%Duin, R.P.W., 2015. The dissimilarity representation for finding universals from particulars by an anti-essentialist approach. \textit{Pattern Recognition Letters}, [online] 64(C), pp.37–43. https://doi.org/10.1016/j.patrec.2015.04.015.
%
%Gelman, S., 2004. Psychological essentialism in children. \textit{Trends in Cognitive Sciences}, [online] 8(9), pp.404–409. https://doi.org/10.1016/j.tics.2004.07.001.
%
%Gelman, S.A., 2005. \textit{Essentialism in Everyday Thought}. [online] https://www.apa.org. Available at: {\textless}https://www.apa.org/science/about/psa/2005/05/gelman{\textgreater} [Accessed 12 January 2023].
%
%Gibbs, C., 2018. Causal essentialism and the identity of indiscernibles. \textit{Philosophical Studies}, [online] 175(9), pp.2331–2351. https://doi.org/10.1007/s11098-017-0961-y.
%
%Kripke, S.A., 1972. Naming and Necessity. In: D. Davidson and G. Harman, eds. \textit{Semantics of Natural Language}, Synthese Library. [online] Dordrecht: Springer Netherlands. pp.253–355. https://doi.org/10.1007/978-94-010-2557-7\_9.
%
%Krzanowski, R. and Polak, P., 2022a. Ontology and AI Paradigms. \textit{Proceedings}, [online] 81(1), p.119. https://doi.org/10.3390/proceedings2022081119.
%
%Krzanowski, R. and Polak, P., 2022b. The Meta-Ontology of AI systems with Human-Level Intelligence. \textit{Philosophical Problems in Science (Zagadnienia Filozoficzne w~Nauce}), (73), p.??-??
%
%Lapuschkin, S., Wäldchen, S., Binder, A., Montavon, G., Samek, W. and Müller, K.-R., 2019. Unmasking Clever Hans predictors and assessing what machines really learn. \textit{Nature Communications}, [online] 10(1), p.1096. https://doi.org/10.1038/s41467-019-08987-4.
%
%Liu, H. and Motoda, H., 1998. \textit{Feature Selection for Knowledge Discovery and Data Mining}. Norwell, MA: Kluwer Academic Publishers.
%
%Mackie, P., 2006. \textit{How Things Might Have Been: Individuals, Kinds, and Essential Properties}. 1st ed. [online] Oxford: Oxford University Press. https://doi.org/10.1093/0199272204.001.0001.
%
%Marcus, R.B., 1993. \textit{Modalities: Philosophical Essays}. [online] New York: Oxford University Press. Available at: {\textless}http://catdir.loc.gov/catdir/enhancements/fy0638/91048105-t.html{\textgreater} [Accessed 12 January 2023].
%
%Matthews, G.B., 1990. Aristotelian Essentialism. \textit{Philosophy and Phenomenological Research}, [online] 50, pp.251–262. https://doi.org/10.2307/2108042.
%
%Mohri, M., Rostamizadeh, A. and Talwalkar, A., 2018. \textit{Foundations of Machine Learning}. 2nd ed. Adaptive computation and machine learning. [online] Cambridge, MA: The MIT Press. Available at: {\textless}https://cs.nyu.edu/\}~{mohri/mlbook/{\textgreater} [Accessed 13 January 2023].
%
%Palmer, F.R., 1981. \textit{Semantics}. 2nd ed. [online] Cambridge: Cambridge University Press. Available at: {\textless}http://archive.org/details/semantics00palm{\textgreater} [Accessed 13 January 2023].
%
%Pelillo, M., 2013. Introduction: The SIMBAD Project. In: M. Pelillo, ed. \textit{Similarity-Based Pattern Analysis and Recognition}, Advances in Computer Vision and Pattern Recognition. [online] London; Heildelberg; New York; Dordrecht: Springer. pp.1–10. https://doi.org/10.1007/978-1-4471-5628-4\_1.
%
%Pelillo, M. and Scantamburlo, T., 2013. How Mature Is the Field of Machine Learning? In: M. Baldoni, C. Baroglio, G. Boella and R. Micalizio, eds. \textit{AI*IA 2013: Advances in Artificial} \textit{Intelligence}, Lecture Notes in Computer Science. [online] Cham: Springer International Publishing. pp.121–132. https://doi.org/10.1007/978-3-319-03524-6\_11.
%
%Pottier, B., 1964. \textit{Vers une sémantique moderne}. Strassbourg: Klincksieck.
%
%Robertson Ishii, T. and Atkins, P., 2020. Essential vs. Accidental Properties. In: E.N. Zalta, ed. \textit{The Stanford Encyclopedia of Philosophy}, Winter 2020. [online] Stanford, CA: Metaphysics Research Lab, Stanford University. Available at: {\textless}https://plato.stanford.edu/archives/win2020/entries/essential-accidental/{\textgreater}.
%
%Rosch, E.H., 1973. Natural categories. \textit{Cognitive Psychology}, [online] 4(3), pp.328–350. https://doi.org/10.1016/0010-0285(73)90017-0.
%
%Skansi, S., 2018. \textit{Introduction to Deep Learning: From Logical Calculus to Artificial Intelligence}. Undergraduate Topics in Computer Science. [online] Cham: Springer International Publishing. https://doi.org/10.1007/978-3-319-73004-2.
%
%Tunç, B., 2015. Semantics of object representation in machine learning. \textit{Pattern Recognition Letters}, [online] 64(15), pp.30–36. https://doi.org/10.1016/j.patrec.2015.03.016.
%
%Watanabe, S., 1985. \textit{Pattern Recognition: Human and Mechanical}. New York: John Wiley \& Sons, Inc.
%
%Zhang, M., 2015. \textit{Google Photos Tags Two African-Americans As Gorillas Through Facial Recognition Software}. [online] Forbes. Available at: {\textless}https://www.forbes.com/sites/mzhang/2015/07/01/google-photos-tags-two-african-americans-as-gorillas-through-facial-recognition-software/{\textgreater} [Accessed 13 January 2023].
%
%Zheng, A. and Casari, A., 2018. \textit{Feature Engineering for Machine Learning: Principles and Techniques for Data Scientists}. 1st ed. Beijing; Boston; Farnham; Sebastopol; Tokyo: O'Reilly.
%\end{document}

>>>>

\begin{artengenv2auth}{Kristina Šekrst, Sandro Skansi}
	{Machine learning and essentialism}
	{Machine learning and essentialism}
	{Machine learning and essentialism}
	{\textsuperscript{1}Departments of Philosophy and Mathematics, Carnegie Mellon University\\
		\textsuperscript{2}Copernicus Center for Interdisciplinary Studies, Jagiellonian University}
	{\label{sekrst_start}Machine learning and essentialism have been connected in the past by various researchers, in order to state that the main paradigm in machine learning processes is equivalent to choosing the ``essential'' attributes for the machine to search for. Our goal in this paper is to show that there are connections between machine learning and essentialism, but only for some kinds of machine learning, and often not including deep learning methods. Similarity-based approaches, more connected to the overall prototype theory, spanning from psychology and linguistics, seem more suited for pattern recognition and complex deep-learning issues, while for classification problems, mostly for unsupervised learning, essentialism seems like the best choice. In order to illustrate the difference better, we will connect both paths to their sources in other disciplines and see how human psychology influences our decision in machine-learning modeling as well. This leads to a~philosophically very interesting consequence: even in the setting of supervised machine learning, essences are not present in data, but in targets, which in turn means that the categories which purport to be essences are in fact human-made, and hand-coded in the targets. The success of machine learning, therefore, does not give any substantial evidence for the independent existence of essential properties. Our stance here is to state that neither the existence nor the lack of ``essential'' properties in machine learning can lead to metaphysical, i.e., ontological claims.}
	{essentialism, machine learning, accidental properties, similarity-based approach, pattern recognition, modal necessity.}
	{%
%		{\flushright\subbold{Kristina Šekrst}\\\subsubsectit\small{University of Zagreb}\par}%
%		{\flushright\subbold{Sandro Skansi}\\\subsubsectit\small{University of Zagreb}\par}%
				{\flushright\subbold{Kristina Šekrst}\\\subbold{Sandro Skansi}\\\subsubsectit\small{University of Zagreb}\par}%
	}

@@

\section*{Essential and accidental properties: introduction}
\lettrine[loversize=0.13,lines=2,lraise=-0.03,nindent=0em,findent=0.2pt]%
{T}{}he purpose of this paper is to show that the existence of essential-like features in machine learning, or the lack of them, cannot provide an ontological commitment.\footnote{The authors would like to thank the anonymous reviewers for their detailed analyses and insights.} Researchers have connected machine-learning practices with essentialist and anti-essentialist stances, but we feel that such claims ignore that both ``essentialist'' and ``anti-essentialist'' paradigms in machine learning are both influenced by human psychology and have no real consequence on the verification of whether there \textit{are} essential properties in nature or not.\footnote{It is necessary to distinguish between ontological commitments regarding nature and ontological commitments that are necessary in every AI system (we call this difference an ontological gap). The former are the subject of this article, and the latter were analyzed by Krzanowski \& Polak
%\label{ref:RNDU3VddFbPO5}(2022a; 2022b).
\parencites*[][]{krzanowski_ontology_2022}[][]{krzanowski_meta-ontology_2022}.%
}

@@

\section*{Philosophical essentialism}
An \textit{essential} property of an object is a~property that an object must have, while an \textit{accidental} one is the one the object happens could have, but that it could lack. That is, in modal terms,\footnote{Standard modal characterizations have been disseminated with the works of Ruth Barcan Marcus and Saul Kripke. Kripke's work on semantics is taking the truth of a~formula relative to a~possible world, since its truth value depends on what is true in accessible world. See Barcan Marcus
%\label{ref:RNDktVIAPuvSw}(1993)
\parencite*[][]{marcus_modalities_1993} %
 for a~modality synthesis and Kripke 
%\label{ref:RND50iCIBiyUn}(1972)
\parencite*[][]{kripke_naming_1972} %
 for Kripke semantics.} we are talking about \textit{necessity} and \textit{possibility},\footnote{There are, of course, differences between logical and (meta)physical possibilities. Something might not be a~logical contradiction, but still be (meta)physically impossible, i.e., not conforming to the laws of nature, for example, a~man travelling faster than the speed of light. The exact details of such classifications, especially between physical and metaphysical possibilities, are a~matter of debate.} respectively 
%\label{ref:RNDcC0OGIZEmV}(Robertson Ishii and Atkins, 2020):
\parencite[][]{robertson_ishii_essential_2020}: %
 an object is going to possess the essential property in all possible worlds, but for an accidental one, there is a~possible world in which an object lacks such a~property. \textit{Essentialism} is a~standpoint in which (at least) some objects have (at least some) essential properties 
%\label{ref:RNDS25JSRGYR9}(Robertson Ishii and Atkins, 2020).
\parencite[][]{robertson_ishii_essential_2020}. %
 For example, an essential property of Socrates is to have originated from his parents but is not essential for him to have brown hair. An essential property of a~dog is certainly not brown hair since there are dogs of other colors. In philosophy, some essential properties are not a~matter of much debate. For example, a~dog had to have some biological origin. \textit{Canis canis} is also a~dog. But in order to pursue the matter further, there might be various objections to candidates for essential purposes.\footnote{A~concise description of the debate is provided by Cartwright 
%\label{ref:RNDsX991rAklO}(1968, p.615):
\parencite*[][p.615]{cartwright_remarks_1968}: %
 ``What are the essential attributes of, say, Dancer's Image? No doubt it will be counted essential that he is a~horse and accidental that he was disqualified in this year's Kentucky Derby. But what of the attribute of being male, or of being a~thoroughbred, or of not being a~Clydesdale stallion? Here, I~suppose, essentialists may disagree. Indeed, a~reasonable essentialist might well take the position that these are hard cases that admit of no clear decision.}'' Often, a~dog is considered a~``wolf-like descendant'', where various breeds might not conform to this ideal, along with the notion of ``having an upturning tail''.

@@

\section*{Machine-learning basics}
\textit{Machine learning}, as a~part of artificial intelligence and computer science, is a~field of approaches and methods that use data in order to improve their performance on some problems. Artificial intelligence can be seen as a~certain type of philosophical engineering
%\label{ref:RNDwNayypCdyS}(Skansi, 2018, p.vii):
\parencite[][p.vii]{skansi_introduction_2018}: %
 we want ``to build machines that can think, [\ldots] understand the meaning, act rationally, cope with uncertainty, [\ldots] handle and talk about objects''. In a~nutshell, we are replicating standard philosophical concepts. It is no wonder that philosophical concepts are deeply embedded in their methods as well but may seem hidden underneath technical layers.

@@

\section*{Psychological essentialism}
Gelman
%\label{ref:RNDqWgJbFowwn}(2004, p.405)
\parencite*[][p.405]{gelman_psychological_2004} %
 describes how once children learn a~new fact about one member of a~category, they generalize the fact to other members of that category, even if they look substantially different. By four years of age, children display subtlety and flexibility when they make category-based inductive inferences. For Gelman 
%\label{ref:RNDf5jZZXitzC}(2004, p.405),
\parencite*[][p.405]{gelman_psychological_2004}, %
 properties seem to be ``fixed at birth'', demonstrated by the following experiment. A~child might learn about a~newborn kangaroo that was switched at birth, and then went to live with goats. The child was then asked whether the animal would be good at hopping or climbing, or if would it have a~pouch or not. Turns out, preschool children typically reported that it would have been good at hopping and have a~pouch, something that seems \textit{inherent} to kangaroos even for children. Such an understanding seems to appear by about six years of age, and it might be as early as four years of age: the time when children reason about animals, plants, and social categories 
%\label{ref:RNDyFC6Tu5B0p}(Gelman, 2004, p.406).
\parencite[][p.406]{gelman_psychological_2004}.%

@@

\section*{Features and prototypes}
As mentioned, \textit{features} tend to be measurable properties that are successful in discriminating and differentiating between different categories of data. For example, in face detection
%\label{ref:RND7NdRdpncD9}(Bishop, 2006, p.3),
\parencite[][p.3]{bishop_pattern_2006}, %
 we aim to find features that are not only fast to compute but also preserve useful discriminatory information enabling faces to be distinguished from non-faces. The study of feature selection finds its practical needs in machine learning, where a~learning algorithm constructs a~description of a~function from a~set of input/output instances through interaction with the world. Machine learning is more concerned with non-continuous features, while pattern recognition deals with continuous ones. It is not the same, for example, to classify something as a~dog or not, compared to finding a~face or another pattern or blob inside an image 
%\label{ref:RNDjCPrNP8QrO}(Liu and Motoda, 1998, p.2).
\parencite[][p.2]{liu_feature_1998}. %
 Liu and Motoda 
%\label{ref:RNDWGLVGs5aWN}(1998, p.2)
\parencite*[][p.2]{liu_feature_1998} %
 state that many forms of representations for machine-learning functions are available, including first-order logic, which is interesting from a~philosophical standpoint, or weighted networks, but they have focused on features since they are 1) \textit{primitive} 2) \textit{convenient} 3) \textit{independent} 4) \textit{widely used} 5) \textit{reasonably general}, i.e., powerful for many applications.

@@

\section*{Essentialist paradigm(s) in machine learning}
It seems intuitive and obvious that supervised machine learning incorporates some kind of essentialism. That is, we are either given discrete or continuous features in datasets that are used for our predictions, usually whether something is a~member of a~class or not. But there are other kinds of machine learning, and we must not ignore the notion of unsupervised learning. We have already mentioned unsupervised learning, in which a~model tries to establish regularities, clusters, or patterns in previously unseen data. This can be compared to the process of human learning at an early age, in which a~human being tries to generalize the already acquired knowledge. Consider this, even if you are getting an unlabeled dataset of weird alien creatures, you will most certainly be able to connect similar ones together in groups or do classifications, even if you do not know what \textit{is} actually in the background of your dataset. We would like the computer to do the same. For example, if we trained our models on a~certain map, they might recognize landmasses, developed areas, forests, or wetlands and group them together, by finding similarities between them. In non-visual data, you might be presented with some numbers, say bank transfers, and you might connect the usual activity into groups, while the outliers might be suspicious.

@@

\section*{Accidental properties in machine learning}
The processes and disciplines of \textit{feature selection} and \textit{extraction} show us that there is a~strong presupposition that something as essential as a~feature exists. There is no doubt that machine learning today is still enveloped in a~strong essentialist paradigm. In feature engineering,\footnote{For more details about feature engineering, see Zheng and Casari
%\label{ref:RNDzLhpHmjkOb}(2018).
\parencite*[][]{zheng_feature_2018}.%
} a~system automatically discovers representations needed for feature detection. For example, it finds close points (neighbors) in a~graph and clusters data around, say, percentages. If feature engineering is an essentialist stance, what kind of essentialism is it? It seems that is not maximal, but also not minimal, we would expect it to lie somewhere in between, judging by its success factor.

@@

\section*{Machine learning or essentialism?}
Our previous conclusion might imply two separate things. Either there is an anti-cybernetic stance in which human learning that encompasses a~certain kind of innate essentialist knowledge is a~different process in machine learning, or that, for practical purposes, knowledge of essential properties is not a~necessary prerequisite for everyday classifications and predictions. The latter seems more intuitive. It does seem that a~similarity-based approach, mimicking the prototypical relationships found in psychological and linguistic research, may work well in various human and machine usages, along with a~combination of properties (features) together with their relations (cf. feature vectors). For some machine-learning tasks, pure essentialism, often a~binary or Boolean one, works best. We believe that essentialism and anti-essentialisms are not binary choices a~computer scientist or a~philosopher must make in order to describe how processes are being generated and run in machine learning paradigms today, but it is a~matter of choice \textit{for a~specific type of task}. There is no essentialism equated with machine learning, but there is both essentialism and anti-essentialism for specific tasks. For some classification tasks and simple pattern recognitions, essentialist features are often the best choice, and for others, systems will work better with combinations of these properties. For unsupervised learning and pattern recognition, prototypical systems, i.e., similarity-based approaches, perform better. A~philosophical take here is that, at least in machine learning, there is no ontological obligation towards either of these stances, but rightful usage for rightful tasks. The choice of your machine-learning system, and therefore, a~supervised or unsupervised approach, will depend on the type of task in question: \textit{what performs better}. It is just a~matter of technical performance that has no metaphysical consequences of the existence of essentialism or anti-essentialism.

@@

\end{artengenv2auth}
\label{sekrst_stop}
